---
title: "Bird phenology exploration"
author: "Michael W Belitz"
date: '2023-03-07'
output: pdf_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(sjPlot)
knitr::opts_chunk$set(echo = TRUE)

#read in the data
arr <- read.csv("Outputs/birds_hopkinsCorrected.csv") # for greenup values
leps <- read.csv("data/phenoEstimates/adult_bfly_phenometrics_noCountCircles_withFull2020Data.csv") %>% 
  dplyr::rename(cell = HEXcell)
fledge <- readRDS("data/phenoEstimates/MAPS-fledge-dates-2022-02-22.rds") %>% 
  mutate(species = stringr::str_replace(sci_name, pattern = " ",
                                        replacement = "_"))
bird_pc <- readRDS("data/bird_PC_vals.rds")
gdd <- data.table::fread('data/gdd_calcs.txt')
ffp <- read.csv("Outputs/frostFreePeriod_byHex.csv")
# combine phenometrics with gdd & ffp
arr_gdd <- left_join(arr, gdd)
arr_gdd <- left_join(arr_gdd, ffp)
fledge_gdd <- left_join(fledge,gdd)
fledge_gdd <- left_join(fledge_gdd, ffp) %>% 
  filter(!is.na(mean_ffp))
leps_gdd <- left_join(leps, gdd)
leps_gdd <- left_join(leps_gdd, ffp)

#filter to years and cells
greenup <- arr_gdd %>% 
  dplyr::distinct(cell, year, .keep_all = T) %>% 
  dplyr::select(cell, year, gr_mn, spring.gdd, summer.gdd, 
                spring.dev, summer.dev, mean_ffp) %>% 
  dplyr::filter(year >= 2002 & year <= 2017)
greenup$cell <- as.character(greenup$cell)
# scale greenup dataframe
greenup_scaled <- greenup %>% 
  mutate(spring.dev = scale(spring.dev),
         mean_ffp = scale(mean_ffp))

arr_gdd_scaled <- left_join(arr_gdd, bird_pc) %>% 
  mutate(gr_mn = scale(gr_mn),
         spring.dev = scale(spring.dev),
         mean_ffp = scale(mean_ffp))

fledge_gdd <- fledge_gdd %>% 
  filter(!is.na(juv_meanday),
         !is.na(spring.gdd),
         !is.na(spring.dev)) %>% 
  dplyr::select(sci_name, station, year, cell, juv_meanday, 
                spring.gdd, summer.gdd,
                spring.dev, summer.dev, PC1, PC2, species, mean_ffp) %>% 
  filter(year >= 2002 & year <= 2017)

fledge_scaled <- ungroup(fledge_gdd) %>% 
  mutate(spring.dev = scale(spring.dev),
         mean_ffp= scale(mean_ffp)) %>% 
  filter(!is.na(juv_meanday),
         !is.na(spring.gdd),
         !is.na(spring.dev),
         !is.na(year)) 

fledge_scaled$cell <- as.character(fledge_scaled$cell)
fledge_scaled_greenup <- left_join(fledge_scaled, greenup_scaled, by = c("year", "cell"))
#gather mean greenup from 2016-2019
mean_gu <- fledge_scaled_greenup %>% 
  filter(year >= 2016 & year <= 2019) %>% 
  group_by(cell) %>% 
  summarise(benchmark_gu = mean(gr_mn))

fledge_scaled_greenup <- left_join(fledge_scaled_greenup, mean_gu, by = "cell")
fledge_scaled_greenup <- fledge_scaled_greenup %>% 
  mutate(greenup.dev = gr_mn - benchmark_gu)

fledge_scaled_greenup <- mutate(fledge_scaled_greenup, greenup.dev = scale(greenup.dev)) 
fledge_scaled_greenup <- filter(fledge_scaled_greenup, !is.na(greenup.dev))
```

## What is the response of fledge dates to GDD deviation?

We find fledge dates are most sensitive to spring GDD in low latitudes and that birds with high PC1 are most sensitive.
```{r fledge gdd}
tm_fledge <- lmer(juv_meanday ~ spring.dev + mean_ffp +
                   PC1 +
                   spring.dev:mean_ffp +
                   spring.dev:PC1 +
                   (1|station) + (1|sci_name),
                 data = fledge_scaled, na.action = na.fail, REML = F)

summary(tm_fledge)

plot_model(tm_fledge, type = "eff", terms = c("spring.dev", "mean_ffp"))
plot_model(tm_fledge, type = "eff", terms = c("spring.dev", "PC1"))
```

## What is the response of fledge dates to Greenup deviation?
Fledge dates are also most sensitive to greenup in areas with high frost free periods (low-latitudes). No interaction with PC1. These results contradict Youngflesh et al., 2021 findings. They also in part contradict our proposed reasoning in the manuscript that the differences between green-up and GDD may drive our contradictory findings. 

```{r fledge greenup}

tm_fledge2 <- lmer(juv_meanday ~ greenup.dev + mean_ffp.x +
                   PC1 +
                   greenup.dev:mean_ffp.x +
                   greenup.dev:PC1 +
                   (1|station) + (1|sci_name),
                 data = fledge_scaled_greenup, na.action = na.fail, REML = F)

summary(tm_fledge2)
plot_model(tm_fledge2, type = "eff", terms = c("greenup.dev", "mean_ffp.x"))
plot_model(tm_fledge2, type = "eff", terms = c("greenup.dev", "PC1"))
```

## Arrival model
Let's check what arrival looks like given my modeling framework. This should co-align with Youngflesh et al., 2021. It does.

```{r arrival model}
tm_arr <- lmer(arr_GAM_mean ~ spring.dev + mean_ffp +
                   PC1 +
                   spring.dev:mean_ffp +
                   spring.dev:PC1 +
                   (1|cell) + (1|species),
                 data = arr_gdd_scaled, na.action = na.fail, REML = F)

summary(tm_arr)
plot_model(tm_arr, type = "eff", terms = c("spring.dev", "mean_ffp"))
plot_model(tm_arr, type = "eff", terms = c("spring.dev", "PC1"))
```

## These effects are even stronger when looking at greenup as predictor instead of GDD.
We do find the same results as Youngflesh et al., 2021. So arrival is showing the same responses but fledge is doing something different. 

```{r greenup arrival}
tm_arr2 <- lmer(arr_GAM_mean ~ gr_mn + mean_ffp +
                   PC1 +
                   gr_mn:mean_ffp +
                   gr_mn:PC1 +
                   (1|cell) + (1|species),
                 data = arr_gdd_scaled, na.action = na.fail, REML = F)

summary(tm_arr2)
plot_model(tm_arr2, type = "eff", terms = c("gr_mn", "mean_ffp"))
plot_model(tm_arr2, type = "eff", terms = c("gr_mn", "PC1"))

```

## Where does that leave us? 
1) The trait by latitude interactions for the fledge models show the opposite response than is shown in the arrival models.
2) We don't really have evidence that green-up vs GDD interactions differ for fledge models, which we posit as a reason for the differences in-text.
3) Our results also contradict Casey's findings in the phenology-demography manuscript currently in-review that species with lower PC1 values have fledge dates more sensitive to green-up. Why?
  - I don't know but that study occurs at a station-level and we coarsen the GDD values to a cell-level. Our study also is done at a smaller spatial area (eastern U.S. vs. continental U.S.) and thus includes fewer species. 
