---
title: "Lep-Bird Pheno Update"
author: "Michael W Belitz"
date: "2/15/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

```{r phenometrics, echo=FALSE}
## read in phenometrics
birds <- read.csv("Outputs/birds_hopkinsCorrected.csv")
leps <- read.csv("data/phenoEstimates/adult_bfly_phenometrics_noCountCircles_withFull2020Data.csv")
```


## Do phenology deviations within cells and groups of species correlate between adult butterflies and birds?
To do this, I calculated a 5-year mean for each cell/species group using the five most recent years. Next, the deviation was reported as the difference between the phenology estimate and 5-year mean phenology estimate. 

Bird species were grouped into "early", "mid", and "late arriving species based on a hopkins-corrected mean arrival time.

### We find no relationship between the deviation of bird and lep phenology

```{r deviation, echo=FALSE, warning=FALSE, message=FALSE}
bird_bm <- birds %>% 
  dplyr::filter(year >= 2013 & year <= 2017) %>% 
  group_by(cell, season) %>% 
  summarise(bmPheno = mean(arr_GAM_mean),
            bmPheno_SD = sd(arr_GAM_mean),
            uniqueYears = length(unique(year)),
            uniqueSpp = length(unique(species)))

## now leps
lep_bm <- leps %>% 
  dplyr::filter(q50 >= 152 & q50 <=243) %>% 
  dplyr::filter(year >= 2016 & year <= 2020) %>% 
  group_by(HEXcell, code) %>% 
  summarise(bmPheno_5 = mean(q5),
            bmPheno_50 = mean(q50),
            bmPheno_5_SD = sd(q5),
            bmPheno_50_SD = sd(q50),
            uniqueYears = length(unique(year)),
            uniqueSpp = length(unique(code)))

## combine these benchmark values with original values
leps2 <- left_join(leps, lep_bm) %>% 
  dplyr::filter(uniqueYears >= 5) %>% 
  dplyr::mutate(dev_5_lep = q5 - bmPheno_5,
                dev_50_lep = q50 - bmPheno_50,
                sd_5_lep = (q5 - bmPheno_5) / bmPheno_5_SD,
                sd_50_lep = (q50 - bmPheno_50) / bmPheno_50_SD) %>% 
  dplyr::select(HEXcell, year, code, dev_5_lep, dev_50_lep, sd_5_lep, sd_50_lep) %>% 
  dplyr::rename(cell = HEXcell)

## for birds, we must first combine phenometrics
birds2 <- birds %>% 
  group_by(cell, year, season) %>% 
  summarise(mean_arr = mean(arr_GAM_mean),
            sd_arr = sd(arr_GAM_mean))

birds3 <- left_join(birds2, bird_bm) %>% 
  dplyr::filter(uniqueYears >= 5) %>% 
  dplyr::mutate(dev_arr_bird = mean_arr - bmPheno,
                sd_arr_bird = (mean_arr - bmPheno) / sd_arr) %>% 
  dplyr::select(cell, year, season, dev_arr_bird, sd_arr_bird)

## combine bird and lepcells
lep_bird <- left_join(leps2, birds3) %>% 
  dplyr::filter(!is.na(season)) %>% 
  dplyr::filter(dev_5_lep > -50 & dev_5_lep < 50)

ggplot(data = lep_bird) +
  geom_point(mapping = aes(x = dev_5_lep, y = dev_arr_bird)) +
  geom_smooth(mapping = aes(x = dev_5_lep, y = dev_arr_bird), method = "lm") +
  facet_grid(code ~ season) +
  labs(x = "Deviation of adult butterfly emergence", y = "Deviation of mean bird arrival") +
  theme_bw()

```

## Does absolute phenology values show a different story?

### Yes, expected trend shown 
```{r absolute, echo=FALSE, warning=FALSE, message=FALSE}
lep_bird_absolute <- leps %>% 
  dplyr::rename(cell = HEXcell) %>% 
  left_join(., birds2) %>% 
  dplyr::filter(q50 >= 152 & q50 <=243) %>% 
  dplyr::filter(q5 < 200) %>% 
  dplyr::filter(!is.na(season),
                !is.na(mean_arr),
                !is.na(q5))

ggplot(data = lep_bird_absolute) +
  geom_point(mapping = aes(x = q5, y = mean_arr)) +
  geom_smooth(mapping = aes(x = q5, y = mean_arr), method = "lm") +
  facet_grid(code ~ season) +
  labs(x = "Adult butterfly emergence", y = "Mean Bird Arrival") +
  theme_bw()

```

## Where and when are the points falling away from the line?

### Model
bird_arrival ~ lep_onset + (1 | unique_bird-lep_grouping) + (0 + lep_onset | unique_bird-lep_grouping)

### Temporal residuals
```{r temp_resids, echo=FALSE, warning=FALSE, message=FALSE}
library(lme4)
lep_bird_absolute <- lep_bird_absolute %>% 
  mutate(f = paste(code,season,sep = "_"))

m <- lmer(mean_arr ~ q5 + (1 | f) + (0 + q5 | f),
        data = lep_bird_absolute,
        REML = FALSE, lmerControl(optimizer = "bobyqa"))
r <- residuals(m)

lep_bird_absolute_r <- lep_bird_absolute %>% 
  mutate(resid = r)

ggplot(data = lep_bird_absolute_r) +
  geom_point(mapping = aes(x = year, y = resid)) +
  geom_smooth(mapping = aes(x = year, y = resid), method = "lm") +
  theme_light() +
  labs(y = "Residuals") +
  facet_grid(code ~ season)
```

### Spatial residuals
```{r spatial_resids, echo=FALSE, warning=FALSE, message=FALSE}
library(sf)
h <- raster::shapefile("data/hex_grid_crop.shp")

h_sf <- st_as_sf(h) %>% 
  st_transform(crs = "+proj=aea +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
h_sf$cell <- as.integer(h_sf$cell)

h_resids <- left_join(h_sf, lep_bird_absolute_r) %>% 
  dplyr::filter(!is.na(resid))

ggplot(data = h_resids) +
  geom_sf(mapping = aes(fill = resid)) +
  scale_fill_gradient2() +
  theme_classic() +
  facet_grid(code ~ season)
```
